motion Aftereffects in the vestibular system: influence of (spatially congruent) subsequent own-body motion on self-motion perception

### 2017-07-31 manuscript.Rmd
first version of paper and bibliogrpahy

---
title: "Motion Aftereffects in the vestibular system: influence of (spatially congruent) subsequent own-body motion on self-motion perception"
shorttitle: "Reversed motion direction illusion"
author:
  - name: "Roman B. Di Francesco"
    affiliation: "1"
  - name: "Andrew W. Ellis"
    affiliation: "1"
  - name: "Luzia Grabherr"  
    affiliation: "1"
  - name: "Gianluca Macauda"
    affiliation: "1"
    corresponding: yes    # Define only one corresponding author
    address: "Departement of Psychology, University of Bern, Fabrikstrasse 8, 3012 Bern, Switzerland"
    email: "gianluca.macauda@psy.unibe.ch"
  - name: "Fred W. Mast"
    affiliation: "1"
affiliation:
  - id: "1"
    institution: "Departement of Psychology, University of Bern"

@@

class: man
lang: english
figsintext: yes
figurelist: no
tablelist: no
lineno: yes
bibliography:
  - references.bib
output: papaja::apa6_pdf
---

@@

plot_correct <- exp_all_correct %>%
  ggplot(aes(x = soa, y = meanCorrect, shape = cue, color = cue)) + 
  geom_hline(yintercept = .5, linetype = 'dashed') +
  stat_summary(fun.y = mean, geom = 'point') + 
  stat_summary(fun.data = mean_se, geom = "errorbar") + 
  #geom_boxplot() +
  coord_cartesian(ylim = c(0,1), expand = 0) + 
  scale_y_continuous(limits = c(0,1)) + 
  labs(x = 'Inter-Stimulus-Intervall', y = 'Accuracy', 
       title = 'Accuracy over all Conditions', 
       color = 'Motion Congruence', 
       shape = 'Motion Congruence') + 
  ylab('Accuracy') + 
  xlab('Inter-Stimulus-Intervall') +
  facet_grid(~experiment) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(size = 12, face="bold", hjust = 0.5), 
        axis.title = element_text(size = 12, face = 'bold'), 
        legend.title = element_text(size = 12, face = 'bold'), 
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

rt_times_plot <- na.omit(exp_all_rt) %>%
    ggplot(aes(x = soa, y = meanLogRT, shape = cue, color = cue, linetype = cue)) + 
    #stat_summary(fun.y = mean, geom = 'point') + 
    #stat_summary(fun.data = mean_cl_normal, geom = "errorbar") + 
    geom_boxplot() +
    labs(x = 'Inter-Stimulus-Intervall', y = 'Accuracy', 
         title = 'Reaction Times', 
         color = 'Motion Congruence', 
         shape = 'Motion Congruence',
         linetype = 'Motion Congruence') + 
    ylab('Mean Log Reaction Time') + 
    xlab('Inter-Stimulus-Intervall') +
    facet_grid(experiment ~ correct) +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(size = 12, face="bold", hjust = 0.5), 
          axis.title = element_text(size = 12, face = 'bold'), 
          legend.title = element_text(size = 12, face = 'bold'), 
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())

@@

The design of the three experiments with the discrimination tasks is ideal to study the perceptual decision making process. As far as we know, the decision making process has never been investigated in MAEs. However, a joint analysis of response and response time would provide more detailed information about the dynamics of the underlying sensory processing.

A commonly used cognitive model for a joint analysis is the Drift Diffusion Model (DDM) [@ratcliff_diffusion_2008]. In the DDM it is assumed that the decision making process relies on the accumulation of noisy sensory evidence over time. A response is given when a threshold for one of the two possibilities is reached. Several parameters specify the decision making process: The drift rate describes the speed of evidence accumulation for one of the two response options. The starting point indicates a point within the boundary from which evidence accumulation starts. Parameters of interest for the current study are the drift rate and starting point. Other important parameters of the DDM include the boundary separation parameter, which serves as the threshold for the needed evidence to make decision and the non decision time parameter that covers the portion within a response not involved in the decision making process (e.g. motor execution).

Other studies within the field of perceptual decision making [@ellis_vestibular_2017; @wiech_influence_2014] used a DDM to investigate the influence of prior information on different parameters of the decision making process and has recently been proposed as model to study the dynamics of vestibular decision making [@merfeld_dynamics_2016] Applied to MAEs, a first motion stimulation (in our case vestibular stimulation) should influence the judgment of the subsequent motion discrimination task by introducing a bias. However, it remains to be seen whether such a bias shifts the starting point to the opposite direction, which would mean that a first motion alters the perceptual decision making process, or whether the first motion affects the drift rate and thus changes the accumulation process, which would be indicative for altered sensory processing [@wiech_influence_2014]. A combination of both biases is possible as well. Those three possbilities are discussed more extensively in a simplified depicition in Figure \@ref(fig:ddms)).

The goal of this study is twofold: First, we want to replicate the previously found MAE in the vestibular modality. However, here it tested whether a stimulation of the otoliths can evoke MAE for the semicircular canals and vice versa. This would further highlight the cross-modal nature of MAEs. Second and based on the MAE, it is investigated whether the introduced bias for the opposite direction is more likely due to changes in sensory processing or a more general cognitive strategy with the help of a joint analysis of response times and responses in a DDM.

@@

Experiment 1 was originally planed as a modification of the Spatial Cueing paradigm by Posner, Snyder and Davidson [@posner_attention_1980] to examine potential influences of own-body motion stimuli on the perception of subsequent motion stimuli. Similar to Spatial Cueing in the visual system, a cue stimulus was expected to direct covert attention to one side in space.  The spatial cueing effect is usually found for short stimuli with short ISIs [@ruz_review_2002]. Therefore both movements were short (200 ms) and the ISIs ranged from 50 ms to 600 ms.

Participants were first translated in the y-axis (cue) and were then rotated about an earth-vertical axis (target). The first motion was chosen to be stronger compared to the second one to ensure that participants perceive the cue correctly, since this could potentially influence attention. A trial was considered as congruent when both motions had the same spatial direction, e.g. left translations and counter-clockwise rotation. The first motion was chosen to be 3.6 times greater than the intensity of the threshold established in a preliminary study and similar to published data [@valko_vestibular_2012] and the second motion was chosen to be 2.6 times greater than the intensity of the thresholds reported by @grabherr_vestibular_2008.

@@

A six degree of freedom motion base (6DOF2000E, Moog Inc., East Aurora, NY) and in-house software were used to generate the motion stimuli. Cue (antecedent) stimuli consisted of translations with single cycle sinusoidal acceleration and a frequency $f$ of 5 Hz ($a(t) = A\sin(2 \pi ft) = A\sin(\frac{2 \pi t}{T})$, $T = \frac{1}{f}$) along the y-axis (left/right) and the x-axis (forth/back) as used in previous studies [see e.g., @benson_thresholds_1989; @crane_foreaft_2012; @grabherr_vestibular_2008]. Acceleration amplitude (A) was set to $0.25 \frac{m}{s^2}$, i.e. a peak velocity of $0.016 \frac{m}{s}$ ($v_{max} = \frac{AT}{\pi}$) and a displacement of 0.0016 m ($\Delta p = \frac{AT^2}{2\pi}$). For the target (subsequent) stimuli the same acceleration profile as for the cue stimuli was used but with yaw rotations about an earth-vertical axis (left/right). The acceleration amplitude was set to $24 \frac{deg}{s^2}$ ($v_{max} = 1.53 \frac{deg}{s}, \Delta p = .153 deg$).

@@

Within each participant Cue Congruence (three levels: congruent, incongruent and neutral) and inter-stimulus interval (ISI, four levels: 50 ms, 100 ms, 200 ms and 600 ms) were manipulated. In congruent trials y-axis translations were followed by yaw rotations to the corresponding side. In incongruent trials y-axis translations preceded yaw rotations to the opposite side. Neutral trials consisted of a translation along the x-axis and a subsequent yaw rotation. The combination of the two factors resulted in 32 motion sets. Every motion set was repeated six times per participant. In-house software based on labVIEW (National Instruments, Austin, Texas) was used to record participants' response and reaction times.

@@

For all experiments analyses and figures were computed in R [@R-base] using the ggplot2 [@R-ggplot2], brms [@R-brms] and rstan [@R-rstan] packages. Analyses, models and data are freely accessible on the OpenScienceFramework (OSF) (https://osf.io/46nqw/).

To describe data from experiment 1 in more detail, participants' performances were analyzed with a multilevel Bayesian logistic regression model with factors Cue Congruence (congruent, incongruent, neutral) and ISI (50ms, 100ms, 200ms, 600ms). 
<!---To analyze participants' reaction times a multilevel Bayesian regression model with an ex-Gaussian likelihood function and factors Cue Congruence (congruent, incongruent, neutral) and ISI (50ms, 100ms, 200ms, 600ms) was computed. Only correct trials were included as the underlying mechanism leading to a wrong response might differ. Priors were set to the default implemented in brms.-->

To jointly analyze the participants' response and response times, and to study the bias introduced by the first motion, two different hierarchical Drift Diffusion Models [@vandekerckhove_hierarchical_2011] in a Bayesian framework were calculated in brms [@R-brms]. In a hierarchical model group and participant parameters can be estimated at the same time based on the assumption that participants' parameters are drawn from a group distribution [@vandekerckhove_hierarchical_2011; @wiecki_hddm:_2013]. It allows parameter estimates for groups while taking into account variation between participants. Additionally, the Bayesian implementation allows inference of the complete posterior distributions of the parameters of interest. For the first experiment, neutral conditions were left out to simplify the statistical models.

In the second model, for the drift rate fixed effects of the first direction (left/right), the second direction (clockwise or counter-clockwise) were estimated together with the random effects of participants. Moreover, for the starting point the fixed effects of the first motion and ISI along with the random effect of participants was estimated. For the non-decision time the random effects of participants were estimated.

The better model was selected based on the leave-one-out cross-validation (LOO) [@vehtari_practical_2017].

@@

```{r figDescriptiveRT, fig.cap='The mean of the proportion of correct answers in the direction discrimination task for all participants in all conditions and experiments. The means over all participants are represented by red dots for congruent conditions, green triangles for incongruent conditions and the blue squares for neutral condtions (only experiment 1). The error bars represent the Standard Error. The dotted line marks the chance level.',  echo=FALSE, cache = FALSE, warning=FALSE}         
rt_times_plot
```

@@

The logistic regression showed that the motion discrimination performance is below chance when the prior stimulus was in the same direction, e.g. a left translation followed by a counter-clockwise yaw rotation. On the other hand, direction discrimination performance is above chance level when a yaw rotation is preceded by an incongruent or a neutral fore-aft motion (Figure \@ref(fig:figDescriptiveCorrect) - experiment 1). The discrimination performance did not differ meaningfully between the incongruent and neutral motions, as all 95% CIs for the posterior distribution of the different included zero. Overall, the response patterns indicate MAEs, except for ISIs of 600ms.

@@

For the first experiment, the winning model included fixed effects of the first direction (left/right), the second direction (clockwise or counter-clockwise), for the drift rate fixed effects of the first direction (left/right), the second direction (clockwise or counter-clockwise) and the random effects of participants (see figure ddm_exp1). To get an estimate for each condition, no intercept and the triple interaction effect was entered.

This model revealed that evidence was accumulated in the direction of the first motion an ISI of 50ms, if the second motion was to the same direction as the first. It appears that evidence is still accumulated for the direction of the first motion when the second motion is to the opposite, yet those 95% credible intervals include zero. Moreover, the second motion did not meaningfully alter evidence accumulation (all CI include zero, with the median around zero).

For the ISI of 100ms evidence is accumulated for the direction of the first motion, except for the condition of the first motion and second motion to the left, where the 95% CI includes zero. Again, the second motion did not meaningfully alter evidence accumulation (all CI include zero, with the median around zero).
For the ISI of 200ms evidence is accumulated for the direction of the first motion, when the second motion is to the other direction. In the congruent conditions, evidence for the first motion direction is accumulated, yet with the 95% CI barely including zero. Here the second motion alters evidence accumulation meaningfully, i.e. it shifted evidence accumulation towards the correct response of the second motion. For the ISI of 600ms, evidence for the second motion is accumulated, if the second motion is to the right. If the second motion is to the left, evidence for the left is accumulated, however in the congruent condition, the 95% CI includes zero.

```{r ddm1Drift, fig.cap='The means for all conditions and their accuracy',  echo=FALSE, cache = FALSE, warning=FALSE}
better_fig_winning_model_exp1_drift 
```

Moreover, the winning model revealed a slightly shifted starting point to the response "right", if the first motion was to the left and vice versa for an ISI of 50ms. However, the 95% credible interval included zero in both starting points (see figure ddm drift exp 1). For an ISI of 100ms, the starting point was shifted towards the threshold of the right response, if the first motion was to the left, with the 95% credibility interval excluding zero. For the same ISI, the starting point was also slightly shifted to the right threshold, if the first motion was to the right, yet with the 95% credibility interval including zero. For, an ISI of 200 and 600ms the bias was slightly shifted towards the left response if the first motion was to the left and to the right response if the first motion was to the right. However, all 95% credible intervals included zero.

```{r ddm1Bias, fig.cap='The means for all conditions and their accuracy',  echo=FALSE, cache = FALSE, warning=FALSE}
better_fig_winning_model_exp1_bias
```

Overall the winning DDM for experiment 1 revealed mainly an altered sensory processing as evidenced by the drift rates and a tendency towards an altered decision making process prior to the evidence accumulation for short ISIs.

@@

In the first experiment it has been shown that there are cross-modal MAEs within the vestibular system in terms of direction discrimination performance if a yaw rotation is preceded by a  translation in the same spatial direction. Thus, a manipulation of the otoliths apparently influence a sensory decision making process in the semicircular canals. However, MAEs are only found for short pauses between the two motion stimuli. More importantly, in a hierarchical DDM the MAEs appear to be explained by a combination of a variable drift and a shifted starting point to the opposite direction of the first motion. It remains to be seen whether for the incongruent conditions whether the correct responses are due to correct sensory evidence accumulation of the second motion or whether the evidence accumulation process is also biased by the first motion. The same response pattern for neutral trials seems to support the former explanation.

@@

The results of Experiment 1 led to the question whether the consistent directional misperception of the second motion (motion aftereffect) still exists when a rotation (cue) is preceding a translation (target), more concretely whether the cross-modal MAE found in experiment 1  is bidirectional. The sequence of the two motions was therefore interchanged so that a left/right yaw rotation preceded a left/right translation (while the motion trajectories stayed otherwise the same). Only congruent and incongruent cues were presented, since no rotations could serve as “neutral” cues for lateral translations in the horizontal plane as a pitch or roll rotation in an upright participant would lead inevitably to a tilted/skewed subsequent motion.

@@

@@

@@

To jointly analyze the participants' response and response times, and to study the bias introduced by the first motion, two different hierarchical Drift Diffusion Models [@vandekerckhove_hierarchical_2011] were calculated in brms [@R-brms]. Neutral conditions were eliminated from the data set to simplify the statistical models.

In the first model, for the drift rate fixed effects of the first direction (counter-clockwise/clockwise), the second direction were estimated together with the random effects of participants. For the non-decision time and starting point the random effects of participants were estimated.

In the second model, for the drift rate fixed effects of the first direction (counter-clockwise/clockwise), the second direction (left/right) were estimated together with the random effects of participants. Moreover, for the starting point the fixed effects of the first motion and ISI along with the random effect of participants was estimated. For the non-decision time the random effects of participants were estimated.

The better model was selected based on the leave-one-out cross-validation (LOO) [@vehtari_practical_2017].

@@

The logistic regression model (In Fig accuracy plot Experiment 2) shows that the motion discrimination performance is around (for ISIs of 50 and 100 ms) or even above chance level (for ISIs 200 and 600ms) when the prior stimulus was congruent with the motion, e.g. a counter-clockwise yaw rotation followed by a left translation. Direction discrimination performance for the second motion is higher when a yaw rotation is followed by an incongruent  translation. In this Experiment there are no clear MAEs.

### Drift Diffusion Models
For the second experiment, the winning model included fixed effects of the first direction (left/right), the second direction (clockwise or counter-clockwise), for the drift rate fixed effects of the first direction (left/right), the second direction (clockwise or counter-clockwise) and the random effects of participants (see figure ddm_exp1).

This model revealed that evidence was accumulated in the direction of the second motion for an ISI of 50, 100 and 200ms, if the second motion was to the opposite direction. If it was to the same direction, evidence for the second motion was accumulated, however with the 95% CI including zero. same direction as the first. It appears that evidence is still accumulated for the direction of the first motion when the second motion is to the opposite, yet those 95% credible intervals include zero. The second motion meaningfully shifted the evidence accumulation process to direction of the correct response. 
For the ISI of 600ms, evidence for the second motion was always accumulated accumulated. The second motion did not meaningfully alter evidence accumulation (all CI include zero, with the median around zero).
Moreover, the winning model revealed no shift of starting point altered as all 95% CI included zero.

```{r ddm2Drift, fig.cap='The means for all conditions and their accuracy',  echo=FALSE, cache = FALSE, warning=FALSE}
better_fig_winning_model_exp2_drift 
```

```{r ddm2Bias, fig.cap='The means for all conditions and their accuracy',  echo=FALSE, cache = FALSE, warning=FALSE}
better_fig_winning_model_exp2_bias
```

>>>>

---
title: "motion Aftereffects in the vestibular system: influence of (spatially congruent) subsequent own-body motion on self-motion perception"
shorttitle: "Reversed motion direction illusion"
author:
  - name: "Roman B. Di Francesco"
    affiliation: "1"
  - name: "Andrew W. Ellis"
    affiliation: "1"
  - name: "Luzia Grabherr"  
    affiliation: "1"
  - name: "Gianluca Macauda"
    affiliation: "1"
    corresponding: yes    # Define only one corresponding author
    address: "Departement of Psychology, University of Bern, Fabrikstrasse 8, 3012 Bern, Switzerland"
    email: "gianluca.macauda@psy.unibe.ch"
  - name: "Fred W. Mast"
    affiliation: "1"
affiliation:
  - id: "1"
    institution: "Departement of Psychology, University of Bern"

@@

class: man
lang: english
figsintext: yes
figurelist: no
tablelist: no
lineno: yes
bibliography:
  - references.bib
output: papaja::apa6_word
---

@@

plot_correct <- exp_all_correct %>%
  ggplot(aes(x = soa, y = meanCorrect, shape = cue, color = cue)) + 
  geom_hline(yintercept = .5, linetype = 'dashed') +
  stat_summary(fun.y = mean, geom = 'point') + 
  stat_summary(fun.data = mean_se, geom = "errorbar") + 
  #geom_boxplot() +
  coord_cartesian(ylim = c(0,1), expand = 0) + 
  scale_y_continuous(limits = c(0,1)) + 
  labs(x = 'Inter-Stimulus-Intervall', y = 'Accuracy', 
       title = 'Accuracy over all Conditions', 
       color = 'motion Congruence', 
       shape = 'motion Congruence') + 
  ylab('Accuracy') + 
  xlab('Inter-Stimulus-Intervall') +
  facet_grid(~experiment) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(size = 12, face="bold", hjust = 0.5), 
        axis.title = element_text(size = 12, face = 'bold'), 
        legend.title = element_text(size = 12, face = 'bold'), 
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

rt_times_plot <- na.omit(exp_all_rt) %>%
    ggplot(aes(x = soa, y = meanLogRT, shape = cue, color = cue, linetype = cue)) + 
    #stat_summary(fun.y = mean, geom = 'point') + 
    #stat_summary(fun.data = mean_cl_normal, geom = "errorbar") + 
    geom_boxplot() +
    labs(x = 'Inter-Stimulus-Intervall', y = 'Accuracy', 
         title = 'Reaction Times', 
         color = 'motion Congruence', 
         shape = 'motion Congruence',
         linetype = 'motion Congruence') + 
    ylab('Mean Log Reaction Time') + 
    xlab('Inter-Stimulus-Intervall') +
    facet_grid(experiment ~ correct) +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(size = 12, face="bold", hjust = 0.5), 
          axis.title = element_text(size = 12, face = 'bold'), 
          legend.title = element_text(size = 12, face = 'bold'), 
          panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())

exp_all_descriptive_plot_subj <- exp_all_correct %>%
  ggplot(aes(x = soa, y = meanCorrect, shape = cue, color = cue)) + 
  geom_hline(yintercept = .5, linetype = 'dashed') +
  geom_point() + 
  #geom_boxplot() +
  scale_y_continuous(limits = c(0,1)) + 
  labs(x = 'Inter-Stimulus-Intervall', y = 'Accuracy', 
       title = 'Accuracy over all Conditions', 
       color = 'Motion Congruence', 
       shape = 'Motion Congruence') + 
  ylab('Accuracy') + 
  xlab('Inter-Stimulus-Intervall') +
  facet_grid(~experiment) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(size = 12, face="bold", hjust = 0.5), 
        axis.title = element_text(size = 12, face = 'bold'), 
        legend.title = element_text(size = 12, face = 'bold'), 
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

@@

The design of the three experiments with the discrimination tasks is ideal to study the perceptual decision making process. As far as we know, the decision making process has never been investigated in MAEs. However, a joint analysis of response and response time would provide more detailed information about the dynamics of the underlying sensory processing. A commonly used cognitive model for a joint analysis is the Drift Diffusion Model (DDM) [@ratcliff_diffusion_2008]. In the DDM it is assumed that the decision making process relies on the accumulation of noisy sensory evidence over time. A response is given when a threshold for one of the two possibilities is reached. Several parameters specify the decision making process: The drift rate describes the speed of evidence accumulation for one of the two response options. The starting point indicates a point within the boundary from which evidence accumulation starts. Parameters of interest for the current study are the drift rate and starting point. Other important parameters of the DDM include the boundary separation parameter, which serves as the threshold for the needed evidence to make decision and the non decision time parameter that covers the portion within a response not involved in the decision making process (e.g. motor execution).

Other studies within the field of perceptual decision making [@ellis_vestibular_2017; @wiech_influence_2014] used a DDM to investigate the influence of prior information on different parameters of the decision making process and has recently been proposed as model to study the dynamics of vestibular decision making [@merfeld_dynamics_2016] Applied to MAEs, a first motion stimulation (in our case vestibular stimulation) should influence the judgment of the subsequent motion discrimination task by introducing a bias. However, it remains to be seen whether such a bias shifts the starting point to the opposite direction, which would mean that a first motion alters the perceptual decision making process, or whether the first motion affects the drift rate and thus changes the accumulation process, which would be indicative for altered sensory processing [@wiech_influence_2014]. A combination of both biases should be considered as well. Those three possbilities are discussed more extensively in a simplified depicition in Figure \@ref(fig:ddms)).

The goal of this study is twofold: First, we want to replicate the previously found vestibular MAE [@coniglio_human_2014; @crane_foreaft_2012; @crane_roll_2012]. Additionally, here it is tested whether a stimulation of the otoliths can evoke MAEs in the semicircular canals and vice versa. This crossmodal effect is closer to situations in daily life. Second and based on the MAE, it is investigated whether the introduced bias for the opposite direction is more likely due to changes in dynamic sensory processing or related to bias introduced prior to sensory processing with the help of a joint analysis of response times and responses in a DDM.

@@

Experiment 1 was originally planed as a modification of the Spatial Cueing paradigm by Posner, Snyder and Davidson [@posner_attention_1980] to examine potential influences of own-body motion stimuli on the perception of subsequent motion stimuli. Similar to Spatial Cueing in the visual system, a cue stimulus (first motion) was expected to direct covert attention to one side in space.  The spatial cueing effect is usually found for short stimuli with short ISIs [@ruz_review_2002]. Therefore both motions were short (200 ms) and the ISIs ranged from 50 ms to 600 ms.

Participants were first translated in the y-axis (cue) and were then rotated about an earth-vertical axis (target). The first motion was chosen to be stronger compared to the second one to ensure that participants perceive the cue correctly, since this could potentially influence attention. A trial was considered as congruent when both motions had the same spatial direction, e.g. left translations and counter-clockwise rotation. A trial was defined incongruent when the second motion was directed to the opposite spatial direction, e.g. left translations and counter-clockwise rotation. Additionally, neutral trials were used to create a baseline. In those trials the yaw rotation was preceded by a motion along the x-axis (forth/back). The first motion was chosen to be 3.6 times greater than the intensity of the threshold established in a preliminary study and similar to published data [@valko_vestibular_2012] and the second motion was chosen to be 2.6 times greater than the intensity of the thresholds reported by @grabherr_vestibular_2008.

@@

A six degree of freedom motion base (6DOF2000E, Moog Inc., East Aurora, NY) and in-house software were used to generate the motion stimuli. Cue stimuli consisted of translations with single cycle sinusoidal acceleration and a frequency $f$ of 5 Hz ($a(t) = A\sin(2 \pi ft) = A\sin(\frac{2 \pi t}{T})$, $T = \frac{1}{f}$) along the y-axis (left/right) and the x-axis (forth/back) as used in previous studies [see e.g., @benson_thresholds_1989; @crane_foreaft_2012; @grabherr_vestibular_2008]. Acceleration amplitude (A) was set to $0.25 \frac{m}{s^2}$, i.e. a peak velocity of $0.016 \frac{m}{s}$ ($v_{max} = \frac{AT}{\pi}$) and a displacement of 0.0016 m ($\Delta p = \frac{AT^2}{2\pi}$). For the target (subsequent) stimuli the same acceleration profile as for the cue stimuli was chosen but with yaw rotations about an earth-vertical axis (left/right). The acceleration amplitude was set to $24 \frac{deg}{s^2}$ ($v_{max} = 1.53 \frac{deg}{s}, \Delta p = .153 deg$).

@@

Within each participant Cue Congruence (three levels: congruent, incongruent and neutral) and inter-stimulus interval (ISI, four levels: 50 ms, 100 ms, 200 ms and 600 ms) were manipulated. In congruent trials y-axis translations were followed by yaw rotations to the corresponding side. In incongruent trials y-axis translations preceded yaw rotations to the opposite direction. Neutral trials consisted of a translation along the x-axis and a subsequent yaw rotation. The combination of the two factors resulted in 32 motion sets. Every motion set was repeated six times per participant. In-house software based on labVIEW (National Instruments, Austin, Texas) was used to record participants' response and reaction times.

@@

All experiments analyses and figures were computed in R [@R-base] using the ggplot2 [@R-ggplot2], brms [@R-brms] and rstan [@R-rstan] packages. Analyses, models and data are freely accessible on the OpenScienceFramework (OSF) (https://osf.io/46nqw/).

To describe data from experiment 1 in more detail, participants' performances were analyzed with a multilevel Bayesian logistic regression model including treatment coded factors Cue Congruence (congruent, incongruent, neutral) and ISI (50ms, 100ms, 200ms, 600ms). 
<!---To analyze participants' reaction times a multilevel Bayesian regression model with an ex-Gaussian likelihood function and factors Cue Congruence (congruent, incongruent, neutral) and ISI (50ms, 100ms, 200ms, 600ms) was computed. Only correct trials were included as the underlying mechanism leading to a wrong response might differ. Priors were set to the default implemented in brms.-->

To jointly analyze the participants' response and response times, and to study the bias introduced by the first motion, two different hierarchical Drift Diffusion Models [@vandekerckhove_hierarchical_2011] in a Bayesian framework were calculated in brms [@R-brms]. In a hierarchical model group and participant parameters can be estimated at the same time based on the assumption that participants' parameters are drawn from a group distribution [@vandekerckhove_hierarchical_2011; @wiecki_hddm:_2013]. It allows parameter estimates for groups while taking into account variation between participants. Additionally, the Bayesian implementation allows inference of the complete posterior distributions of the parameters of interest. We decided against fitting a model with no pooling as we would not have had enough datapoints to get reliable estimates. For the first experiment, neutral conditions were left out to simplify the statistical models. Moreover, congruent and incongruent conditions were split into their specific direction to get a separate factor for the first and second motion.

In the first model, for the drift rate fixed effects of the first direction, the second direction and the ISI were estimated together with the random effects of participants. For the non-decision time and starting point the random effects of participants were estimated.

In the second model, for the drift rate fixed effects of the first direction (left/right), the second direction (clockwise or counter-clockwise) and the ISI were estimated together with the random effects of participants. Moreover, for the starting point the fixed effects of the first motion and ISI along with the random effect of participants were estimated. For the non-decision time the random effects of participants were calculated.  
The better model was selected based on the leave-one-out cross-validation (LOO) [@vehtari_practical_2017].

@@

@@

The logistic regression showed that the motion discrimination performance was below chance when the prior stimulus was in the same direction, e.g. a left translation followed by a counter-clockwise yaw rotation. On the other hand, direction discrimination performance was above chance level when a yaw rotation is preceded by an incongruent or a neutral fore-aft motion (Figure \@ref(fig:figDescriptiveCorrect) - experiment 1). The discrimination performance did not differ meaningfully between the incongruent and neutral motions, as all 95% Credible Intervals (CI) for the posterior distribution of the difference included zero. Overall, the response patterns indicate MAEs, except for ISIs of 600ms.

@@

For the first experiment, the winning model included fixed effects of the first motion direction, the second motion direction and ISI, and the random effects of participants (see Figure \@ref(fig:ddm1Drift)) for the drift rate as well as fixed effects of the first direction, ISI and random effects of participants for the starting point.

This model revealed that evidence was accumulated in the opposite direction of the first motion for an ISI of 50ms, if both motions were to the left. If the first motion was to the right and the second to the left (incongruent condition) evidence for the direction of the second motion was accumuluated. For the second motion to the right slightly more evidence was accumulated for the right if the first motion was to the left, and for the left if the first motion was to the right. Yet those two 95% CIs include zero.
For the ISI of 100ms evidence is accumulated for the direction of the second motion (however, the 95% CI for left-left included zero), if the first motion was to the opposite direction (however, the 95% CI for left-left included zero), otherwise evidence for the opposite direction is accumulated. except for the condition of the first motion and second motion to the left, where the 95% CI included zero. 
For the ISI of 200ms evidence for the direction of the second motion was accumulated if the first motion had been to the opposite direciton. In the congruent conditions, evidence for the first motion direction was accumulated, however with the 95% CI barely including zero.
For the ISI of 600ms, evidence for the second motion was accumulated, if the second motion had been to the right. If the second motion was to the left, evidence for the left was accumulated, however in the left congruent condition, the 95% CI included zero.

```{r ddm1Drift, fig.cap='The parameter estimates of the drift rate of the winning model for each condition separately in experiment 1. Depicted are the median (dot), the 50 (bold line) and 95% (solid line) credible intervals.',  echo=FALSE, cache = FALSE, warning=FALSE}
better_fig_winning_model_exp1_drift 
```

The winning model revealed a slightly shifted starting point to the response "right", if the first motion was to the left and vice versa for an ISI of 50ms. However, the 95% CI included zero in both starting points (see Figure \@ref(fig:ddm1Bias)). For an ISI of 100ms, the starting point was shifted towards the threshold of the right response, if the first motion was to the left. For the same ISI, the starting point was also slightly shifted to the right threshold, if the first motion was to the right, yet with the 50% and 95% CI including zero. For an ISI of 200 and 600ms the bias was slightly shifted towards the left response if the first motion was to the left and to the right response if the first motion was to the right. However, all 95% CIs included zero.

```{r ddm1Bias, fig.cap='The parameter estimates of the starting of the winning model for each condition separately in experiment 1. Depicted are the median (dot), the 50 (bold line) and 95% (solid line) credible intervals.',  echo=FALSE, cache = FALSE, warning=FALSE}
better_fig_winning_model_exp1_bias
```

Overall the winning DDM for experiment 1 mainly revealed an altered sensory processing as evidenced by the drift rates and a tendency towards an altered decision making process prior to the evidence accumulation for very short ISIs.

@@

In the first experiment it has been shown that there are cross-modal MAEs within the vestibular system in terms of direction discrimination performance if a yaw rotation is preceded by a translation in the same spatial direction. Thus, a manipulation of the otoliths apparently influences a sensory decision making process in the semicircular canals. However, MAEs are only found for short intervals between the two motion stimuli. More importantly, in a hierarchical DDM the MAEs appear to be explained by a combination of a variable drift and a shifted starting point to the opposite direction of the first motion.

@@

The results of Experiment 1 led to the question whether the consistent directional misperception of the second motion (MAE) would still exist if a rotation (cue) is preceding a translation (target), more concretely whether the cross-modal MAE found in experiment 1 is bidirectional. The sequence of the two motions was therefore interchanged so that a left/right yaw rotation preceded a left/right translation (while the motion trajectories stayed otherwise the same). Only congruent and incongruent cues were presented, since no rotation could serve as 'neutral' cue for lateral translations in the horizontal plane as a pitch or roll rotation in an upright participant would lead inevitably to a tilted/skewed subsequent motion.

@@

### motion stimuli

@@

To describe data from experiment 2 in more detail, participants' accuracies were analyzed for each experiment with a multilevel Bayesian logistic regression model including factors Cue Congruence (congruent, incongruent) and ISI (50ms, 100ms, 200ms, 600ms).

@@

To jointly analyze the participants' response and response times, and to study the bias introduced by the first motion, two different hierarchical Drift Diffusion Models [@vandekerckhove_hierarchical_2011] were calculated in brms [@R-brms].

In the first model, for the drift rate fixed effects of the first direction (counter-clockwise/clockwise), the second direction and the ISI were estimated together with the random effects of participants. For the non-decision time and starting point the random effects of participants were estimated.

In the second model, for the drift rate fixed effects of the first direction, the second direction and ISI were estimated together with the random effects of participants. Moreover, for the starting point the fixed effects of the first motion and ISI along with the random effect of participants was estimated. For the non-decision time the random effects of participants were estimated. 
The better model was selected based on the leave-one-out cross-validation (LOO) [@vehtari_practical_2017].

@@

The logistic regression model showed that the motion discrimination performance is around (for ISIs of 50 and 100 ms) or even above (for ISIs 200 and 600ms) chance level when the prior stimulus was spatially congruent with the subsequent motion, e.g. a counter-clockwise yaw rotation followed by a left translation. Direction discrimination performance for the second motion was higher when a yaw rotation is followed by an incongruent translation. In this Experiment there were no clear MAEs (see also Figure \@ref(fig:figDescriptiveCorrect))

### Drift Diffusion Models
For the second experiment, the winning model included fixed effects of the first direction, the second direction and ISI for the drift rate, and the random effects of participants as well as fixed effects of the first direction and ISI together with random effects of participants for the starting point (see Figure \@ref(fig:ddm2Drift)).  
This model revealed that evidence was accumulated in the direction of the second motion for an ISI of 50, 100, 200 and 600ms, if the first motion was to the opposite direction. In congruent conditions, evidence for the correct direction of motion was accumulated, however with the 95% CIs including zero.
For the ISI of 600ms, evidence was always accumulated for the second motion.
Moreover, the winning model revealed no shift of starting point as all 50 or 95% CIs included zero (Figure \@ref(fig:dd2Bias)).

```{r ddm2Drift, fig.cap='The parameter estimates of the drift rate of the winning model for each condition separately in experiment 2. Depicted are the median (dot), the 50 (bold line) and 95% (solid line) credible intervals.',  echo=FALSE, cache = FALSE, warning=FALSE}
better_fig_winning_model_exp2_drift 
```

```{r ddm2Bias, fig.cap='The parameter estimates of the starting point of the winning model for each condition separately in experiment 2. Depicted are the median (dot), the 50 (bold line) and 95% (solid line) credible intervals.',  echo=FALSE, cache = FALSE, warning=FALSE}
better_fig_winning_model_exp2_bias
```
